--- 
layout: post 
title:  Accounting for exposure days 
published: true 
comments: true
tags: [statistics, R, ecology] 
bibliography: references.bib
---

Every year I suggest a student use Terry Shaffer's log exposure models for nest survival [-@shaffer2004unified]. And every year I spend hours trying to figure out why the code in the help section of `?family` doesn't work. So this year I'm writing it down.

Irregular data collection intervals are an annoying feature of much ecological data. Nest survival data in particular. Try monitoring the fate of dozens or hundreds of nests scattered across a large landscape and make sure you always visit every one every 3 days. There is always going to be slop[^allthecode]. Ignoring that slop leads to biased estimates of survival. What we really want is the daily survival rate $$p$$, but when the intervals vary what we are observing is $$p^d$$ where $$d$$ is the number of days in the interval.   

Harold Mayfield recognized this problem [@mayfield1975suggestions, and references therein],  and his estimator is the classic way to deal with this problem. Coming up with alternatives and improvements to the Mayfield estimator is a cottage industry among statisticians! Terry Schaffer recognized that a simple change to the link function in a binomial glm model achieved the same goal, and gives us access to many powerful modeling tools. For example, Max Post van der Burg [-@burg2010finding] used the modified link function together with generalized additive models to identify periods of increased risk of nest parasitism. 
```{r setup, echo=FALSE, include=FALSE}
# load necessary packages here
```

Terry used SAS in his paper, but the good R fairies provided code for a log exposure day link function in the examples for `family()`. Except the code doesn't work.

```{r, error=TRUE, message=FALSE}
logexp_bad <- function(days = 1)
{
  linkfun <- function(mu) qlogis(mu^(1/days))
  linkinv <- function(eta) plogis(eta)^days
  mu.eta <- function(eta) days * plogis(eta)^(days-1) * binomial()$mu_eta # this is the error
  valideta <- function(eta) TRUE
  link <- paste0("logexp(", days, ")")
  structure(list(linkfun = linkfun, linkinv = linkinv,
                 mu.eta = mu.eta, valideta = valideta, name = link),
            class = "link-glm")
}


# simulate some data with different exposure days
library(tidyverse)
set.seed(30894879)
abc <- data.frame(x=rnorm(100),
           days=sample(1:5, 100, replace=TRUE))
abc <- mutate(abc,
              p=plogis(x),
              p_days=p^days,
              y = rbinom(100, 1, p_days))

glm(y~x, data=abc, family=binomial(link=logexp_bad(days=abc$days))) # fails
```

The problem is subtle, because it doesn't cause anything to fail in an obvious way, even when you try to use it. The definition of the function `mu.eta()` includes a call to `binomial()` to extract the `mu_eta` component, except *it doesn't exist* so `mu.eta()` ends up NULL. The correct component name is `mu.eta`, so:

```{r}
logexp <- function(days = 1)
{
  linkfun <- function(mu) qlogis(mu^(1/days))
  linkinv <- function(eta) plogis(eta)^days
  mu.eta <- function(eta) days * plogis(eta)^(days-1) * binomial()$mu.eta(eta)
  valideta <- function(eta) TRUE
  link <- paste0("logexp(", days, ")")
  structure(list(linkfun = linkfun, linkinv = linkinv,
                 mu.eta = mu.eta, valideta = valideta, name = link),
            class = "link-glm")
}
summary(glm(y~x, data=abc, family=binomial(link=logexp(days=abc$days)))) # works
```

and that works.  

[^allthecode]: All the code for this post, including that not shown, [can be found here](https://github.com/atyre2/atyre2.github.io/raw/master/_drafts/custom_link.Rmd).

# References