---
layout: post
title:  Happy houR!
date: `r Sys.time()`
published: true
tags: [R, model selection, TIL]
---

It's happy houR. I'm in a happy place, and I'm going to spend an hour trying 
to learn something new. Earlier today I watched a video of [Hadley Wickham explaining his approach to handling many models](http://edinbr.org/edinbr/2016/05/11/may-Hadley-Update2-PostingTalk.html). 
Now I want to see if this combination of `purrr` and `broom` can make my approach to multi-model inference easier. 

So here's what I'd have done last year. This is point count data of 3 species of prairie songbirds together with 3 habitat covariates. I'm 
pretty sure this comes from Andrea Hanson's 2007 MSc. thesis, Conservation and beneficial functions of grassland birds in agroecosystems. Normally I would do a bunch of model checking on my most complex model, but I'm in a rush to try `broom`, so I create a list of possible models. With 3 main effects and their 3 interactions, we have 48 possible models to consider. That is far too many for such a limited dataset. Background knowledge suggests that VOR will be important, so all models I consider include that effect. Then I'll add each of the landscape variables in turn, together with the interaction with VOR. 

```{r oldStyle}
Abundances <- read.csv("_data/Abundances.csv")
Abundances <- Abundances[,1:7] # drop bogus excel columns
models = list(DICK~1,DICK~vor, 
              DICK~vor + pc1,
              DICK~vor + pc1 + vor:pc1,
              DICK~vor + pc2,
              DICK~vor + pc2 + vor:pc2,
              DICK~vor + pc1 + pc2,
              DICK~vor + pc1 + pc2 + vor:pc1,
              DICK~vor + pc1 + pc2 + vor:pc2,
              DICK~vor + pc1 + pc2 + vor:pc1 + vor:pc2,
              DICK~(vor + pc1 + pc2)^2)
fits = lapply(models,glm,data=Abundances,family="poisson")
```

Now that I have a list of fitted models, I can get a model selection table:

```{r modelSelection}
library(AICcmodavg)
aictab(fits,c.hat=c_hat(fits[[11]]),modname=as.character(models))
```

Now in Hadley's approach, I would put the formulas and the models as rows in a data.frame. 

```{r}
mods <- data.frame(models = models)

```

Ah. There I seem to be stuck. I'd thought I'd be able to put the list of models into a column of a data.frame. 
I mean, Hadley put a whole column of data.frames into a data.frame! Surely this isn't any more difficult. Can this be a vector? 

```{r}
data.frame(poo=as.vector(models))
```

No. Hmmm. Happy houR is rapidly coming to a close and I haven't achieved my goal. Maybe the trick is to use `tidyr::nest()`? No, because that only worked to put subsets of the variables into single rows. I guess I could coerce the whole thing to a character vector

```{r}
mods <- data.frame(models = as.character(models))
```

OK that works. But now I'll have to coerce that to a formula before fitting ... I guess I can use a function to handle all that

```{r}
fitMods <- function(f){
  glm(as.formula(f), data=Abundances, family = "poisson")
}
map(mods, fitMods)
```

Well nuts. I'm guessing `map()` isn't doing what I'm expecting, which is walking across the rows. But then again ...
```{r}
mods[1,1]
fitMods(mods[1,1])
```

... it's a factor. oh. 

```{r}
mods <- data.frame(models = as.character(models), stringsAsFactors = FALSE)
mods[1,1]
fitMods(mods[1,1])
```

Aha! progress.

```{r}
result <- map(mods$models, fitMods)
class(result)
```

but that's just a list ... OK. I'm calling it a day. Nothing's ever as simple as it seems. 