---
layout: post
title:  Scaling and Centering
date: `r Sys.time()`
published: true
tags: [R]
---

Centering and rescaling covariates is a common task prior to building
almost any sort of statistical model. Although function `scale()` will scale
scale and center numeric matrices, it always returns a `matrix`. Most model
fitting functions take data.frames. And although `scale()` will take a
data.frame as an input, it fails with an error if there is even one column
that is a categorical variable. That's always bugged me, and I've been musing
about a solution for awhile. I've also wanted to try making an R package using
the fancy automation tools available in RStudio and package `devtools`. Today
those musings and desires collided, and I give you [scaler](http://github.com/atyre2/scaler)!

```{r install-instr, eval = FALSE}
install.packages("devtools")
devtools::install_github("atyre2/scaler")
```

The functions in this package provide type-stable
scaling and/or centering while ignoring
non-numeric columns. Like `scale()`, the constants used are stored as
attributes in the result, enabling automatic undoing of these operations.
This is handy for making plots after fitting the models. That bit isn't 
implemented yet, so don't get excited.

My primary reason for putting this package together is to learn the process of 
making an R package. But secondarily, the scaling and unscaling of covariates
is a common source of errors among students in NRES 803 Ecological Statistics. 
I hope that these functions will reduce those errors and make everyone's lives
easier!

I aim for these functions to be fast and to play well with pipes. Remains to be
seen if those goals are met. 

There are a few other alternatives to `scale()` out there. Steven Walker [wrote 
a blog post](https://stevencarlislewalker.wordpress.com/2012/09/24/a-simpler-version-of-the-r-scale-command/)
describing a function that scales a matrix but uses the entire matrix to 
calculate the centering and scaling constants. Package [arm](https://cran.r-project.org/web/packages/arm/index.html)
has `rescale()`, which takes a single vector and scales by 2 times the standard
deviation. This is [Andrew Gelman's recommendation](http://andrewgelman.com/2006/06/21/standardizing_r/), 
because the coefficient then refers to a change of +/- one standard devation
from the mean. `rescale()` also applies to binary categorical variables, but
not to categorical variatles with more levels. I'm very interested in 
scaling/centering of categorical variables, but haven't implemented any of 
those ideas yet. For the moment `scale_df()` and `center_df()` just skip 
non-numeric columns.

Neither `rescale()` nor `simple.scale()` work directly on data.frames. They don't
stash the information used to scale/center anywhere convienent, so undoing 
the scaling means redoing the calculations or storing that information somewhere 
else. Here's an example of the kind of problem that causes for students. 

```{r}
library(scaler)
library(dplyr) # for %>%
library(ggplot2)

sc_iris <- scale_df(iris) %>%
  center_df()
head(sc_iris)

petals_lm <- lm(Petal.Length ~ Petal.Width, data = sc_iris)

# now make a nice plot with predicted values
nd <- data.frame(Petal.Width = seq(0.1,2.5,0.1))
nd$Petal.Length <- predict(petals_lm, newdata = nd)

ggplot(iris, aes(x = Petal.Width, y = Petal.Length)) + 
  geom_point() + 
  geom_path(data=nd)
```

Did you catch the error? The problem is that the figure is made on the scale
of the original data, but the model is fit using the scaled and centered data. 
What I need to do is scale the new data for prediction by the same constants as
the original data, and then unscale the results so the line goes through the
points. I could make the new data using the range of scaled values, but I'd 
still want to unscale them to put the line on the figure. I haven't written the
unscaling or rescaling code yet, so that has to wait for another blog post! 

If you have any thoughts about what `scaler` should (or shouldn't!) do send me
an email, tweet, or PR. 
