---
layout: post
title:  The value of sexing chickens
date: `r Sys.time()`
published: true
tags: [R, Outreach, Simulation, Mixed Models]
---

The other day [I got a bit carried away 
with an answer to a simple question.]({% 
post_url 2016-4-26-the-sex-ratio-of-chicken
s %}) In the post I said buying 
"straight-run" chickens was cheaper than 
buying sexed chickens, which is why knowing
the sex ratio is important to figuring out 
how many to buy. [Someone on twitter said 
they would never buy straight run for that 
reason.](https://twitter.com/TrashBirdEcol/status/725141100449681409)
Which got me wondering. What is the value
of information for a sexed chick?<sup id="a1">[1](#f1)</sup>


```{r, table2, echo=FALSE, include=FALSE}
table2 <- data.frame(Hen = 1:11,
                     Males = c(4,4,113,15,18,9,14,9,11,15,15),
                     Females = c(14,7,7,13,12,9,19,6,9,12,18))

library(lme4)
ind.model <- glmer(cbind(Males,Females)~1 + (1|Hen), data = table2, family=binomial)
summary(ind.model)
```

We have to choose between two options: order
*n* chicks 'straight run', which means 
without sexing, or *m* sexed individuals.
[Assorted brown egg layers](http://www.mypetchicken.com/catalog/Day-Old-Baby-Chicks/Assorted-Brown-Egg-Layers-Standard-p581.aspx) 
are $2.50 each "Straight run" and $3.00 each
as sexed females. I start by assume that a
hen is worth 1 point of utility up to 4 
hens, and roosters are valueless. 

```{r optimize1, eval=TRUE}
utility <- function(h, maxhens = 4){
  u <- ifelse(h < maxhens+1, h, maxhens)
  return(u)
}

mean_p <- ind.model@beta
var_p <- as.numeric(vcov(ind.model)) # uncertainty in p
var_ind <- ind.model@theta^2
logodds <- rnorm(hens, mean = mean_p, sd = sqrt(var_p + var_ind))
# convert to probability
p <-  1 / (1 + exp(-logodds))
males <- rbinom(hens, size = clutchsize, p = p)

library(ggplot2)
binomial_probs$freq <- binomial_probs$d * hens
gg <- ggplot(as.data.frame(males),aes(x=males)) + geom_bar() + geom_line(data=binomial_probs, mapping=aes(x=hens, y=freq))
gg
```

So by including the between hen variation we end up with `r floor(100 * sum(males<4)/hens)` % of orders having more than 4 females. This is about the same as without the between individual variation. However, there is a greater chance of getting 1 or 0 hens (7 or 8 males) in the order.

Not sure what the takehome is here. Randomness is two faced?

<b id="f1">1</b> Yes it does appear I have too much time on my hands. OTOH, I need to figure out VOI anyway.[↩](#a1)

[hays1945]: http://www.jstor.org/stable/2458081 "The primary sex ratio in domestic chickens"