--- 
layout: post 
title:  What's the best way to implement a logisitic model in R?
published: true
comments: true
tags: [keyword1, keyword2] 
---

I'm teaching my population dynamics class using R for the first time. I want the students to use a simple logistic population model to make predictions about how population size will respond to different management actions. So, I need to figure out the best way to implement a discrete time, logisitic growth model in R.

I have students think through the logistic model with graphs of per capita birth and death rates. I don't expect them to be able to write a function for the model from scratch, so I'm going to give them the code. I'm putting it in a function instead of just doing a for loop in a script to facilitate using functional programming tools (e.g. `purrr`). Here is the function I've come up with. If anyone is aware of an implementation of this model I'd love to hear about it![^allthecode]

```{r setup, echo=FALSE, include=FALSE}
# load necessary packages here
library(tidyverse)
```


We will use the form of the logistic that uses explicit birth and death rates:

$$
N_{t+1} = N_t + (b_0 + b_1 N_t) N_t - (d_0 + d_1 N_t) N_t
$$

This model will have 5 parameters: $b_0, b_1, d_0, d_1, N_0$ as well as vector of times.

```{r logfundef, echo=TRUE}
logisticpop <- function(N0, t = 1, b_0 = 1.25,
                        b_1 = -0.01,
                        d_0 = 0.5,
                        d_1 = 0.005){
  N <- vector("numeric", length = length(t))
  N[1] <- N0  # first entry in vector is initial population size
  last_t <- length(N)
  # these next lines will cause warnings if
  # b is not either length 1 or last_t
  if (length(b_0) != last_t) b_0 <- rep(b_0, times=last_t)
  if (length(b_1) != last_t) b_1 <- rep(b_1, times=last_t)
  if (length(d_0) != last_t) d_0 <- rep(d_0, times=last_t)
  if (length(d_1) != last_t) d_1 <- rep(d_1, times=last_t)
  # Now we "loop" and calculate N for each time
  # notice that the code is *exactly* like the equation
  for (i in seq_along(t[-last_t])){
    N[i+1] <- N[i] * (1 + (b_0[i] + b_1[i]*N[i]) -
                        (d_0[i] + d_1[i]*N[i]))
  }
  return(data.frame(t=t, b_0 = b_0,
                    b_1 = b_1,
                    d_0 = d_0,
                    d_1 = d_1,
                    N = N))
}

```

```{r}
M1 <- logisticpop(1, t = 0:20)
ggplot() + 
  geom_line(aes(x=t, y=N), data = M1)
```

This structure gives me the ability to predict the effects of management actions by providing a vector of birth or death rate parameters. For example, if habitat quality starts to degrade at t = 10, then the intercept of the per capita birth rate could be decreasing.

```{r}
b_0 <- c(rep(1.25, times=10),seq(1.2, by=-0.025, length=11))
M2 <- logisticpop(1, t = 0:20, b_0 = b_0)
ggplot() + 
  geom_line(aes(x=t, y=N), data = M1) + 
  geom_line(aes(x=t, y=N), data = M2, color="red")
```

If I want a stochastic model with environmental stochasticity, I simply provide a vector drawn from random numbers.

```{r}
b_0 <- rnorm(21, 1.25, 0.2)
M3 <- logisticpop(1, t = 0:20, b_0 = b_0)
ggplot() + 
  geom_line(aes(x=t, y=N), data = M1) + 
  geom_line(aes(x=t, y=N), data = M3, color="red")
```

Now I want 20 replicates of that model. This is one place where functional programming comes in. I'll make a list of b_0 vectors, then use `map_df()` to call my model with each of those vectors and return the results as a dataframe.

```{r}
inputs <- map(rep(21, times=20), rnorm, mean=1.25, sd=0.2)
M4 <- map_df(inputs, ~logisticpop(N0 = 1, t = 0:20, b_0 = .x), .id = "rep")
ggplot() + 
  geom_line(aes(x=t, y=N, group = rep), data = M4, color="red") +
  geom_line(aes(x=t, y=N), data = M1, size=2) 
```

So that's my best shot. So far it seems to be working for students to give them the function and have the manipulate the parameters. But lot's of time left in the semester. Their next assignment is to use a similar function to predict population growth in their flour beetle (*Tribolium confusum*) populations. 

[^allthecode]: All the code for this post, including that not shown, [can be found here](https://github.com/atyre2/atyre2.github.io/raw/master/_drafts/whats-the-best-logistic-model.Rmd).